DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems with PDEs

### 2012-11-29 Paper_May_2012/main.tex
Kleinere Aenderungen und ein bis zwei Anmerkungen.

\institute{Christian Goll \at
              Institute of Applied Mathematics, University of Heidelberg
              \\Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany\\
              %Tel.: +49 622-154-6154\\
              \email{cgoll@iwr.uni-heidelberg.de}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              %Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
Winnifried Wollner \at
              Universit\"at Hamburg\\
              Department of Mathematics\\
              Bundesstr. 55, 20146 Hamburg, Germany\\
              \email{winnifried.wollner@math.uni-hamburg.de}   
}

@@

This article will provide information on who the {\em Differential 
Equations and Optimization Environment} \dope{}  \cite{dope} 
solves these problems by providing a well structured interface for high-level 
routines integrated
with tools to solve PDEs. At the present stage we provide an interface to 
the finite element library \deal{} and a collection of algorithms to solve
both pure PDE problems as well as optimization problems 
constrained by a PDE. 
This provides an extension to the 
very few tools available doing so; 
successful in-house code examples are Gascoigne \cite{gascoigne}
and RoDoBo \cite{rodobo}. 
Like \dope{}, both have been initiated as well in 
the Numerical Analysis Group in Heidelberg.

The main feature of \dope is to give a unified interface to high level 
algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed in such a way that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.

While \deal{} --which at present is the only FE-toolkit to which we provide an 
interface-- leaves the implementation of all high-level algorithms to the user, 
\dope{} is user-focused by delivering
prefabricated tools which requires adjustments by the user only for parts
connected to his specific problem. The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.

@@

The article is organized as follows. In the main Section
\ref{detailed_description}, a detailed description of 
the main features is given. First, we first explain the solution 
of stationary PDEs, in Section~\ref{subsubsec:stationary problems}, nonstationary PDEs, in Section~\ref{sec:timedep}, and optimization problems, in 
Section~\ref{sec:opt}. 
In all three cases we first described how the interface is designed, and why 
it is designed as such. Then the implementation required by the user is 
demonstrated at an example.
Finally, additional features are discussed in Section~\ref{sec:additional}. 
In the final Section
\ref{applications}, we give three different numerical 
examples which illustrate the versatile features of \dope{}.

@@

Next, we create some quadrature rules (for the evaluation of integrals over
cells \texttt{quad} and faces \texttt{face\_quad}) and a finite element \text{fe}
appropriate for the PDE at hand. We
choose a Gaussian quadrature of a certain degree and the previously mentioned
Taylor-Hood element. These routines are quite standard in each finite element
software and thus we can use those provided by the FE-library \deal{}.

@@

\subsubsection{\texttt{func.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{pde.h}. 
The main difference, and common to all functionals, is that as we only want 
to evaluate a functional, i.e. a number, and not a residual-vector. 
Hence, we loop over the quadrature points and not over the local 
degrees of freedom.

@@

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l$. Let $I:=[0,T]$ be a time interval with end time point value $T$.
The weak for of such a problem then consists of finding some $u$ such that
\[
\int_I \overline{a}(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a spatial domain $\Omega\in\mathbb{R}^d$. The weak for is then to find 
$u = (v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R})}.
\]
We note that the only difference is the appearance of an additional
time integration, the initial-value, and the term 
\[
\int_I (\partial_t v, \phi)\,dt
\]
so that we can reuse our stationary implementation for all the other 
terms if we rescale them appropriately in the time integration.

>>>>

\institute{Christian Goll \at
              University of Heidelberg, Institute of Applied Mathematics, 
              \\Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany\\
              %Tel.: +49 622-154-6154\\
              \email{christian.goll@iwr.uni-heidelberg.de}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              %Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
Winnifried Wollner \at
              Universit\"at Hamburg\\
              Department of Mathematics\\
              Bundesstr. 55, 20146 Hamburg, Germany\\
              \email{winnifried.wollner@math.uni-hamburg.de}   
}

@@

This article will provide information on how the {\em Differential 
Equations and Optimization Environment} \dope{}  \cite{dope} 
solves these problems by providing a well structured interface for high-level 
routines integrated
with tools to solve PDEs. At the present stage we provide an interface to 
the finite element library \deal{} and a collection of algorithms to solve
both pure PDE problems as well as optimization problems 
constrained by a PDE. 
This provides an extension to the 
very few tools available doing so; 
successful in-house code examples are Gascoigne \cite{gascoigne}
and RoDoBo \cite{rodobo}. 
Like \dope{}, both have been initiated as well in 
the Numerical Analysis Group in Heidelberg.

The main feature of \dope{} is to give a unified interface to high level 
algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed in such a way that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.

While \deal{} --which at present is the only FE-toolkit to which we provide an 
interface-- leaves the implementation of all high-level algorithms to the user, 
\dope{} is user-focused by delivering
prefabricated tools which requires\todocg{require? Ist plural} adjustments by the user only for parts
connected to his specific problem. The possible solution of 
optimization problems with PDE constraints is an innovative feature in \dope{}.

@@

The article is organized as follows. In the main Section
\ref{detailed_description}, a detailed description of 
the main features is given. First, we explain the solution 
of stationary PDEs, in Section~\ref{subsubsec:stationary problems}, nonstationary PDEs, in Section~\ref{sec:timedep}, and optimization problems, in 
Section~\ref{sec:opt}. 
In all three cases we first described how the interface is designed, and why 
it is designed as such. Then the implementation required by the user is 
demonstrated at an example.
Finally, additional features are discussed in Section~\ref{sec:additional}. 
In the final Section
\ref{applications}, we give three different numerical 
examples which illustrate the versatile features of \dope{}.

@@

Next, we create some quadrature rules (for the evaluation of integrals over
cells \texttt{quad} and faces \texttt{face\_quad}) and a finite element \texttt{fe}
appropriate for the PDE at hand. We
choose a Gaussian quadrature of a certain degree and the previously mentioned
Taylor-Hood element. These routines are quite standard in each finite element
software and thus we can use those provided by the FE-library \deal{}.

@@

\subsubsection{\texttt{func.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{pde.h}. 
The main difference, and common to all functionals, is that as we only want 
to evaluate a functional, i.e. a number, and not a residual-vector, we loop over the quadrature points and not over the local 
degrees of freedom.

@@

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l$\todocg{Nochmal: Warum wird $l$ eingefuehrt, wenn unten doch wieder die l2-represenation $f$ verwandt wird?}. Let $I:=[0,T]$ be a time interval with end time point value $T$.
The weak \todocg{formulation?} of such a problem then consists of finding some $u$ such that
\[
\int_I \overline{a}(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a spatial domain $\Omega\in\mathbb{R}^d$. The weak \todocg{formulation?} is then to find 
$u = (v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \set{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R})}.
\]
We note that the only difference is the appearance of an additional
time integration, the initial-value, and the term 
\[
\int_I (\partial_t v, \phi)\,dt
\]
so that we can reuse our stationary implementation for all the other 
terms if we rescale them appropriately in the time integration.
